### 2024-03-10 outputs/paper/paper.qmd
simulation and sketches updated

---
title: "Title"
subtitle: "Subtitle"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: https://github.com/iJustinn/Alberta_Mortality_Rate.git"
date: today
date-format: long
abstract: "Abstract"
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

```{r}
#| message: false
#| echo: false

# load packages
library(tidyverse)
library(ggplot2)
library(janitor)
library(dplyr)
library(readr)
library(knitr)
library(kableExtra)
library(reshape2)
```

```{r}
#| message: false
#| echo: false

# read in data
file_path <- "../../inputs/data/raw_data.csv"
data <- read_csv(file_path, show_col_types = FALSE)
```

# Introduction

Regret, defined as an emotional response to the memory of past decisions or actions that one wishes had been different, significantly impacts our daily lives [@citeRegret]. This paper delves into the factors that influence the intensity of regret, with a specific focus on social norms and past behaviors. Understanding these elements is essential for a clearer grasp of the decision-making process and the emotions that result from it.

We categorize social norms into two main types: descriptive norms, which concern beliefs about common behaviors within a group, and injunctive norms, which relate to beliefs about what behaviors are approved or disapproved by others [@citeUNICEF]. Our research involves two experiments aimed at assessing the effect of these norms on regret. The first experiment looks into the role of these norms in the context of deciding whether or not to pick up a hitchhiker, while the second addresses the consequences of selecting an unconventional driving route.

The findings from these experiments reveal that unusual or unexpected events tend to provoke stronger feelings of regret than routine behaviors. However, the influence of both injunctive and descriptive social norms on the level of regret was found to be less significant. These results highlight how deviations from usual behavior, particularly when in conflict with social norms, can intensify feelings of regret. This insight is crucial for a better understanding of the cognitive processes involved in evaluating decisions and the emotional aftermath.

The organization of this paper is methodical and clear. We start with an overview of the theoretical framework concerning social norms and their connection to regret in the introduction. The methodology and findings from our experiments are detailed in the subsequent data section. The discussion section then addresses the implications of our results, outlines the limitations of the study, and suggests directions for future research.

# Data {#sec-data}

Data used in this paper was cleaned and processed with the programming language R [@citeR]. Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `dplyr` [@citeDplyr], `readr` [@citeReadr], `knitr` [@citeknitr], 'kableExtra' [@citeKableExtra], 'reshape2' [@citeReshape2].

## Source

The data utilized in this study replicates experimental findings from the research documented in @citeOriginalPaper. This original research itself is a replication of three seminal experiments designed to explore the influence of past behaviors on regret. The initial two experiments were pioneered by Kahneman and Miller in 1986, with the third experiment conducted by Miller and McFarland in the same year. Furthermore, the original study expands upon these foundational experiments. For the first experiment, it introduces an additional parameter to measure the overall emotional reaction of participants, considering the impact of both types of social norms. In the second experiment, a "Luck" measure was added to determine if the way people estimate probabilities influences their sense of what is typical. The expansion of the third experiment is not covered in this paper, as our examination is limited to the initial two experiments.

## Method
In the first experiment, participants were presented with a well-known scenario by Kahneman and Miller, often referred to as the hitchhiker scenario: "Mr. Jones almost never takes hitch-hikers in his car. Yesterday he gave a man a ride and was robbed. Mr. Smith frequently takes hitch-hikers in his car. Yesterday he gave a man a ride and was robbed" (@citeOriginalPaper). The original study included three comprehension questions to confirm participants' understanding of the scenario, which we will not detail here. To assess the influence of injunctive and descriptive social norms, participants were asked: "Whose behavior do you think is more common in society? Mr.Jones(1)/Mr.Smith(2)" and "Whose behavior do you think will be more criticized by others in society? Mr.Jones(1)/Mr.Smith(2)" respectively. And the level of regret is also assessed by asking the question "Who do you expect will experience greater regret over the episode? Mr.Jones(1)/Mr.Smith(2)" (@citeOriginalPaper). The additional measure of negative effect is measured by asking the question "Contemplating your previous answers about this scenario and factoring in both Mr.Jones and Mr.Smith personal routines and your perceptions of social norms and possible social criticism, who do you think overall experienced more negative feelings about the decision to take a hitch-hiker that day? Mr.Jones(1)/Mr.Smith(2)".

The second experiment involved another scenario by Kahneman and Miller, known as the car accident scenario: "Mr. Adams was involved in an accident when driving home after work on his regular route. Mr. White was involved in a similar accident when driving on a route that he only takes when he wants a change of scenery" (@citeOriginalPaper). As with the first experiment, comprehension questions were included but are not discussed here. The feeling of regret is assessed by asking the question "Who is more upset over the accident? Mr.Adams(1)/Mr.White(2)". The "Luck" measure is examined by the question "Which of the two do you think is less lucky? Mr.Adams(1)/Mr.White(2)".

## Attibutes
The survey dataset initially contained 52 variables. After excluding variables related to the third experiment and comprehension questions, the dataset was narrowed down to 6 key variables for analysis. For the first experiment, 'Sc1_regret' represents participant responses to the regret question. 'Sc1_socnorms1' and 'Sc1_socnorms2' correspond to the assessment of injunctive and descriptive social norms, respectively. 'Sc1_combinednorms' is an additional variable introduced to measure the overall negative emotional response. As mentioned above, 1 in the dataset represents the Mr.Jones, where 2 represents Mr.Smith.

In relation to the second experiment, 'Sc2_regret' denotes the responses to the regret question. The 'Sc2_lucky' variable is an added measure corresponding to participants' perceptions of luck as introduced in the "Luck" question. And 1 stands for Mr.Adams, where 2 is Mr.White.

The top ten rows of filtered data with variables renamed is shown in @tbl-filtereddata.

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: tbl-filtereddata
#| tbl-cap: top 10 rows of filtered data

example_data <- data %>%
  select(Sc1_regret, sc1_socnorms1, sc1_socnorms2, sc1_combinednorms, Sc2_regret, Sc2_lucky) %>%
  rename(
    "Experiment 1 regret" = Sc1_regret,
    "Injuctive norms" = sc1_socnorms1,
    "Descriptive norms" = sc1_socnorms2,
    "Negative effect" = sc1_combinednorms,
    "Experiment 2 regret" = Sc2_regret,
    "Luck" = Sc2_lucky
  ) %>%
  head(10)  # Selecting only the top 10 rows for the table

# Generating the table using kable and kableExtra for styling
example_kable <- kable(example_data, format = "latex", booktabs = TRUE, align='c') %>% # align the text in the middle
  kable_styling(full_width = FALSE, font_size = 7) 
example_kable
```

# Results
## Experiment 1

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: tbl-experiment1
#| tbl-cap: Result of experiment 1

# Filter the data to drop NA values
data_filtered_1 <- data %>%
  filter(!is.na(Sc1_regret) & !is.na(sc1_socnorms1) & !is.na(sc1_socnorms2) & !is.na(sc1_combinednorms))

# Calculate the total counts and counts for Mr.Jones (1) and Mr.Smith (2)
total_counts_1 <- nrow(data_filtered_1)
mr_jones_counts <- sum(data_filtered_1$Sc1_regret == 1)
mr_smith_counts <- total_counts_1 - mr_jones_counts

# Calculate the counts and percentages for Mr.Jones
mr_jones_data <- data.frame(
  Group = "Mr.Jones(exception)",
  Regret = paste(mr_jones_counts, "(", round((mr_jones_counts / total_counts_1) * 100, 1), "%)", sep = ""),
  Injunctive = paste(sum(data_filtered_1$sc1_socnorms1 == 1), "(", round((sum(data_filtered_1$sc1_socnorms1 == 1) / total_counts_1) * 100, 1), "%)", sep = ""),
  Descriptive = paste(sum(data_filtered_1$sc1_socnorms2 == 1), "(", round((sum(data_filtered_1$sc1_socnorms2 == 1) / total_counts_1) * 100, 1), "%)", sep = ""),
  `Negative Effect` = paste(sum(data_filtered_1$sc1_combinednorms == 1), "(", round((sum(data_filtered_1$sc1_combinednorms == 1) / total_counts_1) * 100, 1), "%)", sep = "")
)

# Calculate the counts and percentages for Mr.Smith
mr_smith_data <- data.frame(
  Group = "Mr.Smith(routine)",
  Regret = paste(mr_smith_counts, "(", round((mr_smith_counts / total_counts_1) * 100, 1), "%)", sep = ""),
  Injunctive = paste(sum(data_filtered_1$sc1_socnorms1 == 2), "(", round((sum(data_filtered_1$sc1_socnorms1 == 2) / total_counts_1) * 100, 1), "%)", sep = ""),
  Descriptive = paste(sum(data_filtered_1$sc1_socnorms2 == 2), "(", round((sum(data_filtered_1$sc1_socnorms2 == 2) / total_counts_1) * 100, 1), "%)", sep = ""),
  `Negative Effect` = paste(sum(data_filtered_1$sc1_combinednorms == 2), "(", round((sum(data_filtered_1$sc1_combinednorms == 2) / total_counts_1) * 100, 1), "%)", sep = "")
)

# Combine the data for Mr. Jones and Mr. Smith into one dataframe
combined_data <- rbind(mr_jones_data, mr_smith_data)

# Now use kable to create the table with the combined data
kable_output <- kable(combined_data, "latex", booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, font_size = 7)

kable_output
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: fig-experiment1bar
#| fig-cap: Bar plot of experiment 1

smith_regret_p <- round((mr_smith_counts / total_counts_1) * 100, 1)
smith_injunctive_p <- round((sum(data_filtered_1$sc1_socnorms1 == 2) / total_counts_1) * 100, 1)
smith_descriptive_p <- round((sum(data_filtered_1$sc1_socnorms2 == 2) / total_counts_1) * 100, 1)
smith_negative_p <- round((sum(data_filtered_1$sc1_combinednorms == 2) / total_counts_1) * 100, 1)
jones_regret_p <- round((mr_jones_counts / total_counts_1) * 100, 1)
jones_injunctive_p <- round((sum(data_filtered_1$sc1_socnorms1 == 1) / total_counts_1) * 100, 1)
jones_descriptive_p <- round((sum(data_filtered_1$sc1_socnorms2 == 1) / total_counts_1) * 100, 1)
jones_negative_p <- round((sum(data_filtered_1$sc1_combinednorms == 1) / total_counts_1) * 100, 1)

# Create a data frame with the provided data
df <- data.frame(
  Category = c("Regret", "Injunctive Social Norm", "Descriptive Social Norm", "Negative Affect"),
  Smith_Routine = c(smith_regret_p, smith_injunctive_p, smith_descriptive_p, smith_negative_p),
  Jones_Exception = c(jones_regret_p, jones_injunctive_p, jones_descriptive_p, jones_negative_p)
)

# Convert the data from wide to long format
data_long <- melt(df, id.vars = "Category", variable.name = "Condition", value.name = "Percentage")

# Reorder the factor levels of Category so that 'Regret' comes first
data_long$Category <- factor(data_long$Category, levels = c("Regret", "Injunctive Social Norm", "Descriptive Social Norm", "Negative Affect"), 
                             labels = c("Regret", "Social Norm\nInjunctive", 
                                        "Social Norm\nDescriptive", "Negative\nAffect"))

# Create the bar plot with the categories reordered
ggplot(data_long, aes(x = Category, y = Percentage, fill = Condition)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  scale_fill_manual(values = c("Smith_Routine" = "grey", "Jones_Exception" = "#505050"), 
                    labels = c("Mr.Smith(Routine)", "Mr.Jones(Exception)")) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  theme_minimal() +
  labs(x = "", y = "Percentage", fill = "Condition") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, colour = "black",size = 9), 
    legend.position = "top", legend.title = element_blank())
```

@tbl-experiment1 and @fig-experiment1bar display results from experiment 1. Mr. Jones, who performs an action that is an exception to his routine, reports a higher regret level of 92.1%. Most participants rated Mr.Jones' behavior to be more common (descriptive) as he never gave ride to hitchhiker (95.3) and therefore to be less criticized (injunctive) by the society (9.4%). Mr.Jones is thought to experience more negative feelings about the decision made that day combining previous answers to the two social norms (92.7%). Mr.Smith, associated with routine behavior, shows markedly lower levels in these aspects, participants rated 7.9%, 4.7%, 90.6% and 7.3% for regret, descriptive social norm, injunctive social norm, and negative effect respectively. Indicating that although Mr.Smith is thought to be criticized more in the society, he would experience less regret and negative effect overall.

The disparity in responses suggests that individuals who engage in behaviors that are not typical for them (Mr. Jones) are more likely to experience and be perceived as having stronger regret and more negative emotional responses. The consistency in this pattern across different measures indicates a robust finding, aligning with psychological theories that predict increased emotional consequences for actions that deviate from one's own or society's expectations. The bar plot @fig-experiment1bar quantitatively illustrates these findings. Mr. Jones's experiences with regret and negative affect are visually represented by significantly taller bars compared to Mr. Smith's, reflecting the numerical data.

## Experiment 2

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: tbl-experiment2
#| tbl-cap: Result of experiment 2

# Filter the data to drop NA values
data_filtered_2 <- data %>%
  filter(!is.na(Sc2_regret) & !is.na(Sc2_random_1) & !is.na(Sc2_random_2) & !is.na(Sc2_lucky))

# Calculate the total counts and counts for Mr.Adams (1) and Mr.White (2)
total_counts_2 <- nrow(data_filtered_2)
mr_adams_counts <- sum(data_filtered_2$Sc2_regret == 1)
mr_white_counts <- total_counts_2 - mr_adams_counts

# Calculate the counts and percentages for Mr.Adams
mr_adams_data <- data.frame(
  Group = "Mr.Adams(routine)",
  Regret = paste(mr_adams_counts, "(", round((mr_adams_counts / total_counts_2) * 100, 1), "%)", sep = ""),
  Luck = paste(sum(data_filtered_2$Sc2_lucky == 1), "(", round((sum(data_filtered_2$Sc2_lucky == 1) / total_counts_2) * 100, 1), "%)", sep = "")
)

# Calculate the counts and percentages for Mr. White
mr_white_data <- data.frame(
  Group = "Mr.White(exception)",
  Regret = paste(mr_white_counts, "(", round((mr_white_counts / total_counts_2) * 100, 1), "%)", sep = ""),
  Luck = paste(sum(data_filtered_2$Sc2_lucky == 2), "(", round((sum(data_filtered_2$Sc2_lucky == 2) / total_counts_2) * 100, 1), "%)", sep = "")
)

# Combine the data for Mr. Jones and Mr. Smith into one dataframe
combined_data <- rbind(mr_adams_data, mr_white_data)

# Now use kable to create the table with the combined data
kable_output <- kable(combined_data, "latex", booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, font_size = 7)

kable_output
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: fig-experiment2bar
#| fig-cap: Bar plot of experiment 2

adams_regret_p <- round((mr_adams_counts / total_counts_2) * 100, 1)
adams_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 1) / total_counts_2) * 100, 1)
white_regret_p <- round((sum(data_filtered_2$Sc2_regret == 2) / total_counts_2) * 100, 1)
white_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 2) / total_counts_2) * 100, 1)

# Create a data frame with the provided data
data <- data.frame(
  Category = c("Regret", "Luck"),
  Adams_Routine = c(adams_regret_p, adams_luck_p),
  White_Exception = c(white_regret_p, white_luck_p)
)

# Convert the data from wide to long format
data_long <- melt(data, id.vars = "Category", variable.name = "Condition", value.name = "Percentage")

# Reorder the factor levels of Category so that 'Regret' comes first
data_long$Category <- factor(data_long$Category, levels = c("Regret", "Luck"))

# Create the bar plot with the categories reordered
ggplot(data_long, aes(x = Category, y = Percentage, fill = Condition)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  scale_fill_manual(values = c("Adams_Routine" = "grey", "White_Exception" = "#505050"), 
                    labels = c("Mr.Adams(Routine)", "Mr.White(Exception)")) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) + # label the y-axis every 20%
  theme_minimal() +
  labs(x = "", y = "Percentage", fill = "Condition") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, colour = "black",size = 9), 
    legend.position = "top", legend.title = element_blank())
```

@tbl-experiment2 and @fig-experiment2bar illustrate the outcomes of experiment 2, which addresses perceptions of regret and luck in decision-making under varying conditions of routine and exception.

Mr. Adams, who represents routine behavior, exhibits a regret level of 18.8%, suggesting that participants view routine actions as less regrettable. His luck perception stands at 32.9%, indicating that participants may see routine choices as less influenced by luck. On the contrary, Mr. White, who represents an exception to routine, is associated with a markedly higher regret level at 81.2%. This suggests that non-routine actions are perceived as significantly more regrettable. Furthermore, his luck perception is at 67.1%, which could reflect a belief that deviations from routine are more susceptible to chance and less within an individual's control.

@fig-experiment2bar graphically accentuates these findings, with much higher height bars for Mr.White in terms of regret and luck. This visual comparison conveys the statistical differences and underscores the paper's exploration of the psychological impacts of routine versus exceptional decisions. This visualization complements the statistical data by providing an immediate, comparative view of the emotional weight and perceived randomness assigned to each individual's experience.

# Discussion

## Replication and estimand

## Measurement validity

## Ethical Considerations and Bias

## Limitations and Future Research

## Conclusion

\newpage

\newpage

# References

### 2024-03-10 scripts/00-simulate_data.R

@@ -1,0 +1,39 @@
+#### Preamble ####
+# Purpose: 
+# Author: Ziheng Zhong
+# Date: 10 March 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+
+
+
+#### Workspace setup ####
+# setup all libraries
+library(tibble)
+library(ggplot2)
+library(tidyverse)
+
+
+
+#### simulation ####
+simulation_data <-
+  tibble(
+    cause = rep(x = c("Heart", "Stroke", "Diabetes"), each = 6, times = 3), # Repeats each cause for every year, 3 times
+    year = rep(x = 2015:2020, each = 1, times = 9), # Each year repeated for each cause, matching the total repetitions
+    deaths = rnbinom(n = 54, size = 20, prob = 0.1) # Matching the total number of rows
+  )
+
+
+
+#### chart ####
+ggplot(simulation_data, aes(x = year, y = deaths, color = cause)) +
+  geom_point() + # Add points
+  geom_smooth(method = "lm", se = FALSE) + # Add linear model trendline
+  theme_minimal() +
+  labs(title = "Trend of Deaths by Cause",
+       x = "Year",
+       y = "Number of Deaths") +
+  scale_x_continuous(breaks = unique(simulation_data$year))
+
+

### 2024-03-10 scripts/02-data_cleaning.R

@@ -1,0 +1,21 @@
+#### Preamble ####
+# Purpose: 
+# Author: Ziheng Zhong
+# Date: 10 March 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+
+
+
+#### Workspace setup ####
+# setup all libraries
+library(opendatatoronto)
+library(tidyverse)
+library(janitor)
+library(dplyr)
+library(readr)
+
+
+
+#### xxx ####

### 2024-03-10 scripts/03-modeling.R

@@ -1,0 +1,22 @@
+#### Preamble ####
+# Purpose: 
+# Author: Ziheng Zhong
+# Date: 10 March 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+
+
+
+#### Workspace setup ####
+# setup all libraries
+library(opendatatoronto)
+library(tidyverse)
+library(ggplot2)
+library(janitor)
+library(dplyr)
+library(readr)
+
+
+
+#### xxx ####

### 2024-03-10 scripts/01-data_downloading.R
data download script and raw data added

@@ -1,0 +1,42 @@
+#### Preamble ####
+# Purpose: 
+# Author: Ziheng Zhong
+# Date: 10 March 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+
+
+
+#### Workspace setup ####
+# setup all libraries
+library(opendatatoronto)
+library(tidyverse)
+library(janitor)
+library(dplyr)
+library(readr)
+
+
+
+#### download ####
+raw_data <-
+  read_csv(
+    "https://open.alberta.ca/dataset/03339dc5-fb51-4552-97c7-853688fc428d/resource/3e241965-fee3-400e-9652-07cfbf0c0bda/download/deaths-leading-causes.csv",
+    skip = 2,
+    col_types = cols(
+      `Calendar Year` = col_integer(),
+      Cause = col_character(),
+      Ranking = col_integer(),
+      `Total Deaths` = col_integer()
+    )
+  )
+
+
+
+#### Save data ####
+write_csv(
+  x = raw_data,
+  file = "inputs/data/raw_data.csv"
+)
+
+

### 2024-03-10 scripts/01-data_downloading.R
data download script updated

@@ -18,7 +18,7 @@ library(readr)
 
 
 
-#### download ####
+#### download data ####
 raw_data <-
   read_csv(
     "https://open.alberta.ca/dataset/03339dc5-fb51-4552-97c7-853688fc428d/resource/3e241965-fee3-400e-9652-07cfbf0c0bda/download/deaths-leading-causes.csv",
@@ -33,7 +33,7 @@ raw_data <-
 
 
 
-#### Save data ####
+#### save data ####
 write_csv(
   x = raw_data,
   file = "inputs/data/raw_data.csv"

### 2024-03-10 scripts/01-data_downloading.R
data download script updated

@@ -10,11 +10,8 @@
 
 #### Workspace setup ####
 # setup all libraries
-library(opendatatoronto)
-library(tidyverse)
-library(janitor)
-library(dplyr)
 library(readr)
+library(tidyverse)
 
 
 

### 2024-03-10 scripts/02-data_cleaning.R
data cleaning script updated

@@ -10,12 +10,64 @@
 
 #### Workspace setup ####
 # setup all libraries
-library(opendatatoronto)
-library(tidyverse)
+library(knitr)
 library(janitor)
-library(dplyr)
-library(readr)
+library(tidyverse)
 
 
 
-#### xxx ####
+#### load data ####
+raw_data <- read_csv("inputs/data/raw_data.csv")
+
+
+
+#### data cleaning ####
+cleaned_data <- raw_data %>%
+  clean_names() %>%
+  add_count(cause) %>%
+  mutate(cause = str_trunc(cause, 30))
+
+
+
+#### check top 8 major causes in 2019 ####
+cleaned_data |>
+  filter(
+    calendar_year == 2019,
+    ranking <= 8
+  ) |>
+  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>
+  kable(
+    col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
+    align = c("l", "r", "r", "r", "r"),
+    digits = 0, booktabs = TRUE, linesep = ""
+  )
+
+
+
+#### filtering needed data ####
+top_five <-
+  cleaned_data |>
+  filter(
+    calendar_year == 2019,
+  ) |>
+  slice_max(order_by = desc(ranking), n = 5) |>
+  pull(cause)
+
+filtered_data <-
+  cleaned_data |>
+  filter(cause %in% top_five)
+
+
+
+#### save data ####
+write_csv(
+  x = cleaned_data,
+  file = "outputs/data/cleaned_data.csv"
+)
+
+write_csv(
+  x = filtered_data,
+  file = "outputs/data/filtered_data.csv"
+)
+
+

### 2024-03-10 scripts/04-charts.R
charts scripts updated

@@ -1,0 +1,35 @@
+#### Preamble ####
+# Purpose: 
+# Author: Ziheng Zhong
+# Date: 10 March 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+
+
+
+#### Workspace setup ####
+# setup all libraries
+library(knitr)
+library(janitor)
+library(ggplot2)
+library(tidyverse)
+
+
+
+#### load data ####
+filtered_data <- read_csv("outputs/data/raw_data.csv")
+
+
+
+#### generate charts ####
+filtered_data |>
+  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
+  geom_line() +
+  theme_minimal() +
+  scale_color_brewer(palette = "Set1") +
+  labs(x = "Year", y = "Annual number of deaths in Alberta") +
+  facet_wrap(vars(cause), dir = "v", ncol = 1) +
+  theme(legend.position = "none")
+
+

### 2024-03-11 scripts/03-modeling.R
modeling script updated

@@ -10,13 +10,54 @@
 
 #### Workspace setup ####
 # setup all libraries
-library(opendatatoronto)
-library(tidyverse)
-library(ggplot2)
-library(janitor)
-library(dplyr)
-library(readr)
 
 
 
-#### xxx ####
+#### load data ####
+filtered_data <- read_csv("outputs/data/raw_data.csv")
+
+
+
+#### poisson model ####
+cause_of_death_alberta_poisson <-
+  stan_glm(
+    total_deaths ~ cause,
+    data = filtered_data,
+    family = poisson(link = "log"),
+    seed = 123
+  )
+
+
+
+#### negative binomial model ####
+cause_of_death_alberta_neg_binomial <-
+  stan_glm(
+    total_deaths ~ cause,
+    data = filtered_data,
+    family = neg_binomial_2(link = "log"),
+    seed = 123
+  )
+
+
+
+#### modeling results ####
+name_mapping <- 
+  c("causeAll other forms of chronic ..."
+    = "other chronic",
+    "causeMalignant neoplasms of trac..."
+    = "neoplasms",
+    "causeOrganic dementia"
+    = "dementia",
+    "causeOther chronic obstructive p..."
+    = "obstructive pulmonary"
+  )
+
+modelsummary(
+  list(
+    "Poisson" = cause_of_death_alberta_poisson,
+    "Negative binomial" = cause_of_death_alberta_neg_binomial
+  ),
+  coef_map = name_mapping
+)
+
+

### 2024-03-11 scripts/04-charts.R
charts script updated

@@ -32,4 +32,9 @@ filtered_data |>
   facet_wrap(vars(cause), dir = "v", ncol = 1) +
   theme(legend.position = "none")
 
+pp_check(cause_of_death_alberta_poisson) +
+  theme(legend.position = "bottom")
+
+pp_check(cause_of_death_alberta_neg_binomial) +
+  theme(legend.position = "bottom")
 

### 2024-03-12 scripts/03-modeling.R
scripts updated

@@ -14,7 +14,7 @@
 
 
 #### load data ####
-filtered_data <- read_csv("outputs/data/raw_data.csv")
+filtered_data <- read_csv("outputs/data/filtered_data.csv.csv")
 
 
 

### 2024-03-12 scripts/04-charts.R

@@ -18,7 +18,7 @@ library(tidyverse)
 
 
 #### load data ####
-filtered_data <- read_csv("outputs/data/raw_data.csv")
+filtered_data <- read_csv("outputs/data/filtered_data.csv")
 
 
 

### 2024-03-12 outputs/paper/paper.qmd
paper first draft

@@
 # load packages
-library(tidyverse)
-library(ggplot2)
-library(janitor)
-library(dplyr)
-library(readr)
-library(knitr)
-library(kableExtra)
-library(reshape2)
+packages <- c("tidyverse", "boot", "broom.mixed", "collapse", "dataverse", "gutenbergr", "janitor", "knitr", "marginaleffects", "modelsummary", "rstanarm")
+
+# Function to check and install missing packages
+install_and_load <- function(packages) {
+  for (package in packages) {
+    if (!require(package, character.only = TRUE)) {
+      install.packages(package)
+      library(package, character.only = TRUE)
+    }
+  }
+}
+
+# Call the function with the list of packages
+install_and_load(packages)
 ```
@@
 # read in data
-file_path <- "../../inputs/data/raw_data.csv"
-data <- read_csv(file_path, show_col_types = FALSE)
+file_path <- "../../outputs/data/filtered_data.csv"
+filtered_data <- read_csv(file_path, show_col_types = FALSE)
 ```
@@
 
-Regret, defined as an emotional response to the memory of past decisions or actions that one wishes had been different, significantly impacts our daily lives [@citeRegret].
-This paper delves into the factors that influence the intensity of regret, with a specific focus on social norms and past behaviors.
-Understanding these elements is essential for a clearer grasp of the decision-making process and the emotions that result from it.
+Alberta, a province known for its vibrant culture and dynamic economy, faces ongoing challenges in public health, with mortality rates reflecting a complex interplay of socioeconomic, environmental, and healthcare factors.
+This paper aims to delve into the intricacies of mortality rates in Alberta, exploring the trends, causes, and determinants that shape the province's public health landscape.
+By leveraging data on causes of death, including chronic diseases, accidents, and emerging health threats, we provide a detailed analysis of mortality patterns and their implications for Alberta's healthcare system and policy-making.
 
-We categorize social norms into two main types: descriptive norms, which concern beliefs about common behaviors within a group, and injunctive norms, which relate to beliefs about what behaviors are approved or disapproved by others [@citeUNICEF].
-Our research involves two experiments aimed at assessing the effect of these norms on regret.
-The first experiment looks into the role of these norms in the context of deciding whether or not to pick up a hitchhiker, while the second addresses the consequences of selecting an unconventional driving route.
-
-The findings from these experiments reveal that unusual or unexpected events tend to provoke stronger feelings of regret than routine behaviors.
-However, the influence of both injunctive and descriptive social norms on the level of regret was found to be less significant.
-These results highlight how deviations from usual behavior, particularly when in conflict with social norms, can intensify feelings of regret.
-This insight is crucial for a better understanding of the cognitive processes involved in evaluating decisions and the emotional aftermath.
-
-The organization of this paper is methodical and clear.
-We start with an overview of the theoretical framework concerning social norms and their connection to regret in the introduction.
-The methodology and findings from our experiments are detailed in the subsequent data section.
-The discussion section then addresses the implications of our results, outlines the limitations of the study, and suggests directions for future research.
+This paper utilize data from open.alberta.ca as bases.
 
@@
 
-The data utilized in this study replicates experimental findings from the research documented in @citeOriginalPaper.
-This original research itself is a replication of three seminal experiments designed to explore the influence of past behaviors on regret.
-The initial two experiments were pioneered by Kahneman and Miller in 1986, with the third experiment conducted by Miller and McFarland in the same year.
-Furthermore, the original study expands upon these foundational experiments.
-For the first experiment, it introduces an additional parameter to measure the overall emotional reaction of participants, considering the impact of both types of social norms.
-In the second experiment, a "Luck" measure was added to determine if the way people estimate probabilities influences their sense of what is typical.
-The expansion of the third experiment is not covered in this paper, as our examination is limited to the initial two experiments.
-
 ## Method
-In the first experiment, participants were presented with a well-known scenario by Kahneman and Miller, often referred to as the hitchhiker scenario: "Mr.
-Jones almost never takes hitch-hikers in his car.
-Yesterday he gave a man a ride and was robbed.
-Mr.
-Smith frequently takes hitch-hikers in his car.
-Yesterday he gave a man a ride and was robbed" (@citeOriginalPaper).
-The original study included three comprehension questions to confirm participants' understanding of the scenario, which we will not detail here.
-To assess the influence of injunctive and descriptive social norms, participants were asked: "Whose behavior do you think is more common in society?
-Mr.Jones(1)/Mr.Smith(2)" and "Whose behavior do you think will be more criticized by others in society?
-Mr.Jones(1)/Mr.Smith(2)" respectively.
-And the level of regret is also assessed by asking the question "Who do you expect will experience greater regret over the episode?
-Mr.Jones(1)/Mr.Smith(2)" (@citeOriginalPaper).
-The additional measure of negative effect is measured by asking the question "Contemplating your previous answers about this scenario and factoring in both Mr.Jones and Mr.Smith personal routines and your perceptions of social norms and possible social criticism, who do you think overall experienced more negative feelings about the decision to take a hitch-hiker that day?
-Mr.Jones(1)/Mr.Smith(2)".
-
-The second experiment involved another scenario by Kahneman and Miller, known as the car accident scenario: "Mr.
-Adams was involved in an accident when driving home after work on his regular route.
-Mr.
-White was involved in a similar accident when driving on a route that he only takes when he wants a change of scenery" (@citeOriginalPaper).
-As with the first experiment, comprehension questions were included but are not discussed here.
-The feeling of regret is assessed by asking the question "Who is more upset over the accident?
-Mr.Adams(1)/Mr.White(2)".
-The "Luck" measure is examined by the question "Which of the two do you think is less lucky?
-Mr.Adams(1)/Mr.White(2)".
 
 ## Attibutes
-The survey dataset initially contained 52 variables.
-After excluding variables related to the third experiment and comprehension questions, the dataset was narrowed down to 6 key variables for analysis.
-For the first experiment, 'Sc1_regret' represents participant responses to the regret question.
-'Sc1_socnorms1' and 'Sc1_socnorms2' correspond to the assessment of injunctive and descriptive social norms, respectively.
-'Sc1_combinednorms' is an additional variable introduced to measure the overall negative emotional response.
-As mentioned above, 1 in the dataset represents the Mr.Jones, where 2 represents Mr.Smith.
 
-In relation to the second experiment, 'Sc2_regret' denotes the responses to the regret question.
-The 'Sc2_lucky' variable is an added measure corresponding to participants' perceptions of luck as introduced in the "Luck" question.
-And 1 stands for Mr.Adams, where 2 is Mr.White.
-
-The top ten rows of filtered data with variables renamed is shown in @tbl-filtereddata.
+# Results
 
@@
 #| fig-pos: 'h'
-#| label: tbl-filtereddata
-#| tbl-cap: top 10 rows of filtered data
+#| label: fig-poisson_model
+#| fig-cap: result for poisson model
 
-example_data <- data %>%
-  select(Sc1_regret, sc1_socnorms1, sc1_socnorms2, sc1_combinednorms, Sc2_regret, Sc2_lucky) %>%
-  rename(
-    "Experiment 1 regret" = Sc1_regret,
-    "Injuctive norms" = sc1_socnorms1,
-    "Descriptive norms" = sc1_socnorms2,
-    "Negative effect" = sc1_combinednorms,
-    "Experiment 2 regret" = Sc2_regret,
-    "Luck" = Sc2_lucky
-  ) %>%
-  head(10)  # Selecting only the top 10 rows for the table
-
-# Generating the table using kable and kableExtra for styling
-example_kable <- kable(example_data, format = "latex", booktabs = TRUE, align='c') %>% # align the text in the middle
-  kable_styling(full_width = FALSE, font_size = 7) 
-example_kable
-```
-
-# Results
-## Experiment 1
-
-```{r}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-pos: 'h'
-#| label: tbl-experiment1
-#| tbl-cap: Result of experiment 1
-
-# Filter the data to drop NA values
-data_filtered_1 <- data %>%
-  filter(!is.na(Sc1_regret) & !is.na(sc1_socnorms1) & !is.na(sc1_socnorms2) & !is.na(sc1_combinednorms))
-
-# Calculate the total counts and counts for Mr.Jones (1) and Mr.Smith (2)
-total_counts_1 <- nrow(data_filtered_1)
-mr_jones_counts <- sum(data_filtered_1$Sc1_regret == 1)
-mr_smith_counts <- total_counts_1 - mr_jones_counts
-
-# Calculate the counts and percentages for Mr.Jones
-mr_jones_data <- data.frame(
-  Group = "Mr.Jones(exception)",
-  Regret = paste(mr_jones_counts, "(", round((mr_jones_counts / total_counts_1) * 100, 1), "%)", sep = ""),
-  Injunctive = paste(sum(data_filtered_1$sc1_socnorms1 == 1), "(", round((sum(data_filtered_1$sc1_socnorms1 == 1) / total_counts_1) * 100, 1), "%)", sep = ""),
-  Descriptive = paste(sum(data_filtered_1$sc1_socnorms2 == 1), "(", round((sum(data_filtered_1$sc1_socnorms2 == 1) / total_counts_1) * 100, 1), "%)", sep = ""),
-  `Negative Effect` = paste(sum(data_filtered_1$sc1_combinednorms == 1), "(", round((sum(data_filtered_1$sc1_combinednorms == 1) / total_counts_1) * 100, 1), "%)", sep = "")
+#### poisson model ####
+cause_of_death_alberta_poisson <-
+  stan_glm(
+    total_deaths ~ cause,
+    data = filtered_data,
+    family = poisson(link = "log"),
+    seed = 123
   )
 
-# Calculate the counts and percentages for Mr.Smith
-mr_smith_data <- data.frame(
-  Group = "Mr.Smith(routine)",
-  Regret = paste(mr_smith_counts, "(", round((mr_smith_counts / total_counts_1) * 100, 1), "%)", sep = ""),
-  Injunctive = paste(sum(data_filtered_1$sc1_socnorms1 == 2), "(", round((sum(data_filtered_1$sc1_socnorms1 == 2) / total_counts_1) * 100, 1), "%)", sep = ""),
-  Descriptive = paste(sum(data_filtered_1$sc1_socnorms2 == 2), "(", round((sum(data_filtered_1$sc1_socnorms2 == 2) / total_counts_1) * 100, 1), "%)", sep = ""),
-  `Negative Effect` = paste(sum(data_filtered_1$sc1_combinednorms == 2), "(", round((sum(data_filtered_1$sc1_combinednorms == 2) / total_counts_1) * 100, 1), "%)", sep = "")
+#### negative binomial model ####
+cause_of_death_alberta_neg_binomial <-
+  stan_glm(
+    total_deaths ~ cause,
+    data = filtered_data,
+    family = neg_binomial_2(link = "log"),
+    seed = 123
   )
 
-# Combine the data for Mr. Jones and Mr. Smith into one dataframe
-combined_data <- rbind(mr_jones_data, mr_smith_data)
+filtered_data |>
+  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
+  geom_line() +
+  theme_minimal() +
+  scale_color_brewer(palette = "Set1") +
+  labs(x = "Year", y = "Annual number of deaths in Alberta") +
+  facet_wrap(vars(cause), dir = "v", ncol = 1) +
+  theme(legend.position = "none")
 
-# Now use kable to create the table with the combined data
-kable_output <- kable(combined_data, "latex", booktabs = TRUE) %>%
-  kable_styling(full_width = FALSE, font_size = 7)
+pp_check(cause_of_death_alberta_poisson) +
+  theme(legend.position = "bottom")
 
-kable_output
+pp_check(cause_of_death_alberta_neg_binomial) +
+  theme(legend.position = "bottom")
+
 ```
@@
 #| fig-pos: 'h'
-#| label: fig-experiment1bar
-#| fig-cap: Bar plot of experiment 1
+#| label: tbl-summary
+#| tbl-cap: summary of modeling
 
-smith_regret_p <- round((mr_smith_counts / total_counts_1) * 100, 1)
-smith_injunctive_p <- round((sum(data_filtered_1$sc1_socnorms1 == 2) / total_counts_1) * 100, 1)
-smith_descriptive_p <- round((sum(data_filtered_1$sc1_socnorms2 == 2) / total_counts_1) * 100, 1)
-smith_negative_p <- round((sum(data_filtered_1$sc1_combinednorms == 2) / total_counts_1) * 100, 1)
-jones_regret_p <- round((mr_jones_counts / total_counts_1) * 100, 1)
-jones_injunctive_p <- round((sum(data_filtered_1$sc1_socnorms1 == 1) / total_counts_1) * 100, 1)
-jones_descriptive_p <- round((sum(data_filtered_1$sc1_socnorms2 == 1) / total_counts_1) * 100, 1)
-jones_negative_p <- round((sum(data_filtered_1$sc1_combinednorms == 1) / total_counts_1) * 100, 1)
-
-# Create a data frame with the provided data
-df <- data.frame(
-  Category = c("Regret", "Injunctive Social Norm", "Descriptive Social Norm", "Negative Affect"),
-  Smith_Routine = c(smith_regret_p, smith_injunctive_p, smith_descriptive_p, smith_negative_p),
-  Jones_Exception = c(jones_regret_p, jones_injunctive_p, jones_descriptive_p, jones_negative_p)
+name_mapping <- 
+  c("causeAll other forms of chronic ..."
+    = "other chronic",
+    "causeMalignant neoplasms of trac..."
+    = "neoplasms",
+    "causeOrganic dementia"
+    = "dementia",
+    "causeOther chronic obstructive p..."
+    = "obstructive pulmonary"
   )
 
-# Convert the data from wide to long format
-data_long <- melt(df, id.vars = "Category", variable.name = "Condition", value.name = "Percentage")
-
-# Reorder the factor levels of Category so that 'Regret' comes first
-data_long$Category <- factor(data_long$Category, levels = c("Regret", "Injunctive Social Norm", "Descriptive Social Norm", "Negative Affect"), 
-                             labels = c("Regret", "Social Norm\nInjunctive", 
-                                        "Social Norm\nDescriptive", "Negative\nAffect"))
-
-# Create the bar plot with the categories reordered
-ggplot(data_long, aes(x = Category, y = Percentage, fill = Condition)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
-  scale_fill_manual(values = c("Smith_Routine" = "grey", "Jones_Exception" = "#505050"), 
-                    labels = c("Mr.Smith(Routine)", "Mr.Jones(Exception)")) +
-  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
-  theme_minimal() +
-  labs(x = "", y = "Percentage", fill = "Condition") +
-  theme(axis.text.x = element_text(hjust = 1, vjust = 1, colour = "black",size = 9), 
-    legend.position = "top", legend.title = element_blank())
+modelsummary(
+  list(
+    "Poisson" = cause_of_death_alberta_poisson,
+    "Negative binomial" = cause_of_death_alberta_neg_binomial
+  ),
+  coef_map = name_mapping
+)
 ```
-
-@tbl-experiment1 and @fig-experiment1bar display results from experiment 1.
-Mr.
-Jones, who performs an action that is an exception to his routine, reports a higher regret level of 92.1%.
-Most participants rated Mr.Jones' behavior to be more common (descriptive) as he never gave ride to hitchhiker (95.3) and therefore to be less criticized (injunctive) by the society (9.4%).
-Mr.Jones is thought to experience more negative feelings about the decision made that day combining previous answers to the two social norms (92.7%).
-Mr.Smith, associated with routine behavior, shows markedly lower levels in these aspects, participants rated 7.9%, 4.7%, 90.6% and 7.3% for regret, descriptive social norm, injunctive social norm, and negative effect respectively.
-Indicating that although Mr.Smith is thought to be criticized more in the society, he would experience less regret and negative effect overall.
-
-The disparity in responses suggests that individuals who engage in behaviors that are not typical for them (Mr.
-Jones) are more likely to experience and be perceived as having stronger regret and more negative emotional responses.
-The consistency in this pattern across different measures indicates a robust finding, aligning with psychological theories that predict increased emotional consequences for actions that deviate from one's own or society's expectations.
-The bar plot @fig-experiment1bar quantitatively illustrates these findings.
-Mr.
-Jones's experiences with regret and negative affect are visually represented by significantly taller bars compared to Mr.
-Smith's, reflecting the numerical data.
-
-## Experiment 2
-
-```{r}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-pos: 'h'
-#| label: tbl-experiment2
-#| tbl-cap: Result of experiment 2
-
-# Filter the data to drop NA values
-data_filtered_2 <- data %>%
-  filter(!is.na(Sc2_regret) & !is.na(Sc2_random_1) & !is.na(Sc2_random_2) & !is.na(Sc2_lucky))
-
-# Calculate the total counts and counts for Mr.Adams (1) and Mr.White (2)
-total_counts_2 <- nrow(data_filtered_2)
-mr_adams_counts <- sum(data_filtered_2$Sc2_regret == 1)
-mr_white_counts <- total_counts_2 - mr_adams_counts
-
-# Calculate the counts and percentages for Mr.Adams
-mr_adams_data <- data.frame(
-  Group = "Mr.Adams(routine)",
-  Regret = paste(mr_adams_counts, "(", round((mr_adams_counts / total_counts_2) * 100, 1), "%)", sep = ""),
-  Luck = paste(sum(data_filtered_2$Sc2_lucky == 1), "(", round((sum(data_filtered_2$Sc2_lucky == 1) / total_counts_2) * 100, 1), "%)", sep = "")
-)
-
-# Calculate the counts and percentages for Mr. White
-mr_white_data <- data.frame(
-  Group = "Mr.White(exception)",
-  Regret = paste(mr_white_counts, "(", round((mr_white_counts / total_counts_2) * 100, 1), "%)", sep = ""),
-  Luck = paste(sum(data_filtered_2$Sc2_lucky == 2), "(", round((sum(data_filtered_2$Sc2_lucky == 2) / total_counts_2) * 100, 1), "%)", sep = "")
-)
-
-# Combine the data for Mr. Jones and Mr. Smith into one dataframe
-combined_data <- rbind(mr_adams_data, mr_white_data)
-
-# Now use kable to create the table with the combined data
-kable_output <- kable(combined_data, "latex", booktabs = TRUE) %>%
-  kable_styling(full_width = FALSE, font_size = 7)
-
-kable_output
-```
-
-```{r}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-pos: 'h'
-#| label: fig-experiment2bar
-#| fig-cap: Bar plot of experiment 2
-
-adams_regret_p <- round((mr_adams_counts / total_counts_2) * 100, 1)
-adams_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 1) / total_counts_2) * 100, 1)
-white_regret_p <- round((sum(data_filtered_2$Sc2_regret == 2) / total_counts_2) * 100, 1)
-white_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 2) / total_counts_2) * 100, 1)
-
-# Create a data frame with the provided data
-data <- data.frame(
-  Category = c("Regret", "Luck"),
-  Adams_Routine = c(adams_regret_p, adams_luck_p),
-  White_Exception = c(white_regret_p, white_luck_p)
-)
-
-# Convert the data from wide to long format
-data_long <- melt(data, id.vars = "Category", variable.name = "Condition", value.name = "Percentage")
-
-# Reorder the factor levels of Category so that 'Regret' comes first
-data_long$Category <- factor(data_long$Category, levels = c("Regret", "Luck"))
-
-# Create the bar plot with the categories reordered
-ggplot(data_long, aes(x = Category, y = Percentage, fill = Condition)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
-  scale_fill_manual(values = c("Adams_Routine" = "grey", "White_Exception" = "#505050"), 
-                    labels = c("Mr.Adams(Routine)", "Mr.White(Exception)")) +
-  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) + # label the y-axis every 20%
-  theme_minimal() +
-  labs(x = "", y = "Percentage", fill = "Condition") +
-  theme(axis.text.x = element_text(hjust = 1, vjust = 1, colour = "black",size = 9), 
-    legend.position = "top", legend.title = element_blank())
-```
-
-@tbl-experiment2 and @fig-experiment2bar illustrate the outcomes of experiment 2, which addresses perceptions of regret and luck in decision-making under varying conditions of routine and exception.
-
-Mr.
-Adams, who represents routine behavior, exhibits a regret level of 18.8%, suggesting that participants view routine actions as less regrettable.
-His luck perception stands at 32.9%, indicating that participants may see routine choices as less influenced by luck.
-On the contrary, Mr.
-White, who represents an exception to routine, is associated with a markedly higher regret level at 81.2%.
-This suggests that non-routine actions are perceived as significantly more regrettable.
-Furthermore, his luck perception is at 67.1%, which could reflect a belief that deviations from routine are more susceptible to chance and less within an individual's control.
-
-@fig-experiment2bar graphically accentuates these findings, with much higher height bars for Mr.White in terms of regret and luck.
-This visual comparison conveys the statistical differences and underscores the paper's exploration of the psychological impacts of routine versus exceptional decisions.
-This visualization complements the statistical data by providing an immediate, comparative view of the emotional weight and perceived randomness assigned to each individual's experience.
 
@@
 
-## Replication and estimand
-
-## Measurement validity
-
-## Ethical Considerations and Bias
-
-## Limitations and Future Research
-
-## Conclusion
+# Conclusion
 

### 2024-03-12 outputs/paper/paper.qmd
paper updated

@@
 # read in data
-file_path <- "../../outputs/data/filtered_data.csv"
-filtered_data <- read_csv(file_path, show_col_types = FALSE)
+file_path_0 <- "../../inputs/data/raw_data.csv"
+file_path_1 <- "../../outputs/data/filtered_data.csv"
+file_path_2 <- "../../outputs/data/cleaned_data.csv"
+
+raw_data <- read_csv(file_path_0, show_col_types = FALSE)
+filtered_data <- read_csv(file_path_1, show_col_types = FALSE)
+cleaned_data <- read_csv(file_path_2, show_col_types = FALSE)
 ```
@@
 
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: tbl-causes
+#| tbl-cap: top 8 major causes in 2019
+
+#### check top 8 major causes in 2019 ####
+cleaned_data |>
+  filter(
+    calendar_year == 2019,
+    ranking <= 8
+  ) |>
+  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>
+  kable(
+    col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
+    align = c("l", "r", "r", "r", "r"),
+    digits = 0, booktabs = TRUE, linesep = ""
+  )
+
+```
+
 ## Attibutes
@@
 ```{r}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-pos: 'h'
-#| label: fig-poisson_model
-#| fig-cap: result for poisson model
+#| include: false
 
@@
   )
+
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-poisson_model
+#| fig-cap: result for poisson model
 

### 2024-03-12 scripts/02-data_cleaning.R
scripts libraries updated

@@ -10,6 +10,7 @@
 
 #### Workspace setup ####
 # setup all libraries
+library(readr)
 library(knitr)
 library(janitor)
 library(tidyverse)

### 2024-03-12 scripts/03-modeling.R

@@ -10,7 +10,9 @@
 
 #### Workspace setup ####
 # setup all libraries
-
+library(readr)
+library(rstanarm)
+library(modelsummary)
 
 
 #### load data ####

### 2024-03-12 scripts/04-charts.R

@@ -10,6 +10,7 @@
 
 #### Workspace setup ####
 # setup all libraries
+library(readr)
 library(knitr)
 library(janitor)
 library(ggplot2)

### 2024-03-12 outputs/paper/paper.qmd
paper updated

@@
 
-This paper utilize data from open.alberta.ca as bases.
-
 # Data {#sec-data}
@@
 Data used in this paper was cleaned and processed with the programming language R [@citeR].
-Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `dplyr` [@citeDplyr], `readr` [@citeReadr], `knitr` [@citeknitr], 'kableExtra' [@citeKableExtra], 'reshape2' [@citeReshape2].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary].
 
 ## Source
+
+This paper utilize data from open.alberta.ca [@citeDatasource] as bases.
 

### 2024-03-13 outputs/paper/paper.qmd
Data and Result added

@@
 ---
-title: "Title"
-subtitle: "Subtitle"
+title: "Modeling Mortality Trends in Alberta"
+subtitle: "A Quantitative Approach to Public Health Dynamics"
 author: 
@@
 
-This paper utilize data from open.alberta.ca [@citeDatasource] as bases.
+In the pursuit of understanding and modeling mortality trends within Alberta, our study primarily utilizes the "Leading Causes of Death" dataset made available by the Alberta Open Government Portal.
+This comprehensive data source offers an extensive overview of mortality statistics across the province, categorized by various causes of death.
+The Alberta Open Government Portal, a key initiative for promoting transparency and public engagement, hosts this dataset at open.alberta.ca [@citeDatasource], providing open access to valuable health-related data.
+
+The "Leading Causes of Death" dataset encompasses a detailed compilation of death counts and mortality rates spanning multiple years.
+It includes categorizations by age, sex, and specific causes of death, ranging from chronic diseases to accidents and emergent health threats.
+This granularity allows for a nuanced analysis of mortality trends, facilitating the identification of prevalent health issues and potential public health interventions.
+
+The selection of this data source was driven by its credibility, comprehensiveness, and relevance to our research objectives.
+Being an official government resource, the dataset ensures reliability and accuracy, crucial for the validity of our analytical models.
+The dataset's extensive coverage of demographic and cause-specific mortality information enables a thorough exploration of the factors influencing mortality rates in Alberta.
+Furthermore, the data's organization and accessibility significantly aid in the efficient processing and modeling of mortality trends.
 
@@
 
-## Attibutes
+Our analysis commenced with the essential task of refining the raw dataset.
+The initial phase involved cleaning and standardizing the dataset's variable names to enhance readability and analytical efficiency.
+We also aggregated the data to quantify the prevalence of each cause of death, thereby streamlining the dataset for more focused analysis.
+A critical aspect of our methodology was the isolation of data from the year 2019.
+This strategic choice was made to examine mortality trends immediately before the onset of the COVID-19 pandemic.
+By focusing on this specific year, we aimed to establish a baseline understanding of mortality patterns unaffected by the pandemic's impact.
+Within this scope, as showing in @tbl-causes, we identified and prioritized the top 8 causes of mortality, based on their significance and prevalence.
+
+Building on the refined dataset, our analysis further narrowed to investigate the top 5 causes of mortality in 2019.
+This deliberate focus allowed us to concentrate our efforts on the most impactful health challenges faced by Alberta before the pandemic.
+We applied statistical models to fit these mortality trends, seeking to identify underlying patterns and determinants.
+This approach was aimed at uncovering insights that could inform public health strategies and interventions, contributing to a more nuanced understanding of mortality dynamics in the province.
 
@@
 
+## Data Trend
+
 ```{r}
@@
 #| fig-pos: 'h'
-#| label: fig-poisson_model
-#| fig-cap: result for poisson model
+#| label: fig-result
+#| fig-cap: Trend for Top 5 Causes
 
@@
   facet_wrap(vars(cause), dir = "v", ncol = 1) +
-  theme(legend.position = "none")
-
-pp_check(cause_of_death_alberta_poisson) +
-  theme(legend.position = "bottom")
-
-pp_check(cause_of_death_alberta_neg_binomial) +
-  theme(legend.position = "bottom")
+  theme(legend.position = "none",
+        axis.text.y = element_text(size = rel(0.7)))
 
 ```
+
+Results in the @fig-result displays the annual number of deaths in Alberta for the top five causes in 2019 unveils distinct trends for each cause over approximately two decades.
+The data suggest a relatively stable trend in deaths from acute myocardial infarction, with a small rise that may indicate a steady prevalence of risk factors or an equilibrium between advancing medical interventions and demographic shifts.
+Conversely, the increasing trajectory seen in deaths due to chronic diseases points to a potential escalation in lifestyle-related health issues or could reflect the demographic realities of an aging population.
+
+Particularly notable is the sharp increase in deaths attributed to organic dementia, surging upwards more markedly in the latter half of the time span.
+This trend could signify heightened awareness and improved diagnostic capabilities, alongside the impact of an aging society.
+Meanwhile, deaths from malignant neoplasms related to the respiratory system display minor fluctuations, hinting at persistent challenges in cancer prevention, detection, and treatment.
+The upward trend in chronic obstructive pulmonary disease-related mortality could be linked to environmental factors or historic smoking patterns, underscoring the pressing need for robust public health policies and effective healthcare strategies.
+
+## Modeling
 
@@
 
+Then the @tbl-summary summarizes the coefficients of both models built to fit the mortality trend in Alberta, offering a comparison of their estimations for the top causes of mortality.
+In both models, "other chronic" conditions and "neoplasms" exhibit positive coefficients, suggesting these causes of death have an increasing trend over time.
+Meanwhile, "obstructive pulmonary" diseases show a negative relationship in both models, indicating a decreasing trend in mortality rates.
+The consistency across both models reinforces the reliability of our findings.
+
+As depicted by the graphs for the Poisson (@fig-poisson_model) and negative binomial models (@fig-negative_binomial_model), the diagnostic plots further elucidate the model fits.
+The negative binomial model, appears to capture the dispersion in the data more effectively than the Poisson model, which is evident in two plots.
+The spread of the simulated lines (y_rep) around the observed data (y) suggests that the negative binomial model may accommodate the overdispersion present in the data, which is a common characteristic in count data such as mortality rates.
+
+The presence of multiple peaks in the second plot's simulated data distributions indicates potential subgroups within the data or non-linearity in the mortality trends that the Poisson model may not be accounting for.
+Such patterns underscore the importance of considering different model structures when dealing with complex datasets.
+These graphical diagnostics not only affirm the adequacy of the negative binomial model over the Poisson model for this particular dataset but also highlight the necessity of thorough model checking in the process of statistical analysis.
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-poisson_model
+#| fig-cap: Poisson Model
+
+pp_check(cause_of_death_alberta_poisson) +
+  theme(legend.position = "bottom")
+
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-negative_binomial_model
+#| fig-cap: Negative Binomial Model
+
+pp_check(cause_of_death_alberta_neg_binomial) +
+  theme(legend.position = "bottom")
+
+```
+
 # Discussion

### 2024-03-15 outputs/paper/paper.qmd
data revision, paper result & discussion updated

@@
 # load packages
-packages <- c("tidyverse", "boot", "broom.mixed", "collapse", "dataverse", "gutenbergr", "janitor", "knitr", "marginaleffects", "modelsummary", "rstanarm")
+packages <- c("tidyverse", "boot", "broom.mixed", "collapse", "dataverse", "gutenbergr", "janitor", "knitr", "marginaleffects", "modelsummary", "rstanarm", "tidybayes")
 
@@
 file_path_0 <- "../../inputs/data/raw_data.csv"
-file_path_1 <- "../../outputs/data/filtered_data.csv"
-file_path_2 <- "../../outputs/data/cleaned_data.csv"
+file_path_1 <- "../../outputs/data/cleaned_data.csv"
+file_path_2 <- "../../outputs/data/top_5_cause_data.csv"
+file_path_3 <- "../../outputs/data/summarized_top_5_cause_data.csv"
 
 raw_data <- read_csv(file_path_0, show_col_types = FALSE)
-filtered_data <- read_csv(file_path_1, show_col_types = FALSE)
-cleaned_data <- read_csv(file_path_2, show_col_types = FALSE)
+cleaned_data <- read_csv(file_path_1, show_col_types = FALSE)
+top_5_cause_data <- read_csv(file_path_2, show_col_types = FALSE)
+summarized_top_5_cause_data <- read_csv(file_path_3, show_col_types = FALSE)
 ```
@@
 Data used in this paper was cleaned and processed with the programming language R [@citeR].
-Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary], `tidybayes` [@citeTidybayes].
 
@@
     total_deaths ~ cause,
-    data = filtered_data,
+    data = top_5_cause_data,
     family = poisson(link = "log"),
@@
     total_deaths ~ cause,
-    data = filtered_data,
+    data = top_5_cause_data,
     family = neg_binomial_2(link = "log"),
@@
 #| fig-pos: 'h'
-#| label: fig-result
-#| fig-cap: Trend for Top 5 Causes
+#| label: fig-trends
+#| fig-cap: Trend for Alberta Mortality
+#| fig-subcap: ["Top 5 Causes","Sum of Causes"]
+#| layout-ncol: 2
+#| layout-nrow: 1
 
-filtered_data |>
+top_5_cause_data |>
   ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
@@
   scale_color_brewer(palette = "Set1") +
-  labs(x = "Year", y = "Annual number of deaths in Alberta") +
+  labs(x = "Year", y = "Annual Number of deaths in Alberta") +
   facet_wrap(vars(cause), dir = "v", ncol = 1) +
@@
 
+summarized_top_5_cause_data |>
+  ggplot(aes(x = calendar_year, y = sum_deaths)) +
+  geom_line() +
+  theme_minimal() +
+  scale_x_continuous(breaks = unique(summarized_top_5_cause_data$calendar_year)) +
+  labs(x = "Year", y = "Sum of Annual Number of Deaths in Alberta") +
+  theme(legend.position = "none",
+        axis.text.x = element_text(size = rel(0.7)))
+
 ```
 
-Results in the @fig-result displays the annual number of deaths in Alberta for the top five causes in 2019 unveils distinct trends for each cause over approximately two decades.
+Results in the @fig-trends-1 displays the annual number of deaths in Alberta for the top five causes in 2019 unveils distinct trends for each cause over approximately two decades.
 The data suggest a relatively stable trend in deaths from acute myocardial infarction, with a small rise that may indicate a steady prevalence of risk factors or an equilibrium between advancing medical interventions and demographic shifts.
@@
 The upward trend in chronic obstructive pulmonary disease-related mortality could be linked to environmental factors or historic smoking patterns, underscoring the pressing need for robust public health policies and effective healthcare strategies.
+
+The graph presented in @fig-trends-2 illustrates a clear upward trajectory in the total number of deaths from 2001 to 2022.
+This trend, while indicative of a rising mortality rate, necessitates a nuanced interpretation.
+The increasing death toll may not necessarily be attributed to the lethality or prevalence of specific causes alone.
+Instead, it potentially reflects broader demographic changes, such as the growth of the overall population, particularly within older age groups who are statistically more vulnerable to various health-related fatalities.
 
@@
 
-As depicted by the graphs for the Poisson (@fig-poisson_model) and negative binomial models (@fig-negative_binomial_model), the diagnostic plots further elucidate the model fits.
+For the Poisson model, the expected log count is defined as:
+
+$$
+\log(\mu_i) = \beta_0 + 0.447 \times (\text{other chronic})_i + 0.223 \times (\text{neoplasms})_i + 0.046 \times (\text{dementia})_i -
+$$
+$$
+0.205 \times (\text{obstructive pulmonary})_i
+$$
+
+For the Negative Binomial model, the expected log count is defined as:
+
+$$
+\log(\mu_i) = \beta_0 + 0.449 \times (\text{other chronic})_i + 0.226 \times (\text{neoplasms})_i + 0.048 \times (\text{dementia})_i -
+$$
+$$
+0.203 \times (\text{obstructive pulmonary})_i
+$$
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-models
+#| fig-cap: Model Results
+#| fig-subcap: ["Poisson Model","Negative Binomial Model"]
+#| layout-ncol: 2
+#| layout-nrow: 1
+
+pp_check(cause_of_death_alberta_poisson) +
+  theme(legend.position = "bottom")
+
+pp_check(cause_of_death_alberta_neg_binomial) +
+  theme(legend.position = "bottom")
+
+```
+
+Then to see how good two models fits the original data.
+We can reffer to the values in the lower half of @tbl-summary, but it's always more clear to see the visual, as depicted by the graphs for the Poisson (@fig-models-1) and negative binomial models (@fig-models-2), the diagnostic plots further elucidate the model fits.
 The negative binomial model, appears to capture the dispersion in the data more effectively than the Poisson model, which is evident in two plots.
@@
 
-```{r}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-pos: 'h'
-#| label: fig-poisson_model
-#| fig-cap: Poisson Model
-
-pp_check(cause_of_death_alberta_poisson) +
-  theme(legend.position = "bottom")
-
-```
-
-```{r}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-pos: 'h'
-#| label: fig-negative_binomial_model
-#| fig-cap: Negative Binomial Model
-
-pp_check(cause_of_death_alberta_neg_binomial) +
-  theme(legend.position = "bottom")
-
-```
-
 # Discussion
 
+The preceding sections have laid the groundwork for a comprehensive analysis of mortality trends in Alberta, paving the way for this discussion.
+The application of Poisson and Negative Binomial models to the data has illuminated key trends and has provided quantitative support to the qualitative observations made earlier.
+The positive coefficients for "other chronic" and "neoplasms" in both models suggest that deaths from these causes have steadily increased, raising questions about the broader factors at play beyond the healthcare system's direct influence.
+Simultaneously, the negative coefficient for "obstructive pulmonary" diseases suggests a decline in mortality from these conditions, potentially reflecting successful public health initiatives and changing social habits, such as reduced smoking rates.
+
+However, the increasing overall mortality trend signals a need to consider demographic shifts, such as population growth and aging, as significant contributors to the observed patterns.
+The spike in deaths from 2020 onwards aligns with the onset of the COVID-19 pandemic, underscoring the substantial impact of infectious disease outbreaks on mortality statistics.
+This outlier emphasizes the importance of having robust health surveillance systems that can adapt to sudden changes in public health landscapes.
+
+Furthermore, the diagnostic plots from both models underscore the necessity for careful selection of appropriate statistical methods when modeling complex health data.
+The better fit of the Negative Binomial model compared to the Poisson model, as indicated by the diagnostic checks, suggests that overdispersion—a common feature in count data—must be accounted for to avoid underestimating the variability in the data.
+
+The limitations of the current study also warrant discussion.
+While the models applied provide valuable insights, they do not capture all the nuances of the data.
+Future research could explore the inclusion of additional variables, such as economic indicators, healthcare access measures, and lifestyle factors, to better understand the determinants of mortality.
+Moreover, the investigation of interactions between causes of death and demographic variables could yield a more granular understanding of the dynamics at play.
+
+Ultimately, this study serves as a call to action for ongoing surveillance, research, and a proactive approach to public health planning.
+By continuing to refine our models and expand our datasets, we can better anticipate and mitigate the factors contributing to mortality, striving for a healthier future for all Albertans.
+
 # Conclusion
+
+In conclusion, our exploration of mortality trends in Alberta has provided a statistical narration of the province's public health status over the past two decades.
+We have identified critical causes of death and trends that warrant attention from both healthcare providers and policymakers.
+The analysis has also emphasized the significance of employing appropriate statistical models to accurately represent and understand complex health data.
+As we move forward, it is imperative to integrate these findings into a strategic framework that enhances Alberta's health system responsiveness and effectiveness in the face of demographic and epidemiological shifts.
+Through vigilance and informed action, we can aim to influence these mortality trends positively, enhancing the health and well-being of the Alberta population.
 

### 2024-03-15 scripts/02-data_cleaning.R
scripts updated

@@ -45,7 +45,7 @@ cleaned_data |>
 
 
 
-#### filtering needed data ####
+#### top 5 data ####
 top_five <-
   cleaned_data |>
   filter(
@@ -54,9 +54,17 @@ top_five <-
   slice_max(order_by = desc(ranking), n = 5) |>
   pull(cause)
 
-filtered_data <-
+top_5_cause_data <-
   cleaned_data |>
   filter(cause %in% top_five)
+
+
+
+#### summarize top 5 data ####
+summarized_top_5_cause_data <- top_5_cause_data %>%
+  group_by(calendar_year) %>%
+  summarize(sum_deaths = sum(total_deaths, na.rm = TRUE)) %>%
+  ungroup()
 
 
 
@@ -67,8 +75,12 @@ write_csv(
 )
 
 write_csv(
-  x = filtered_data,
-  file = "outputs/data/filtered_data.csv"
+  x = top_5_cause_data,
+  file = "outputs/data/top_5_cause_data.csv"
 )
 
+write_csv(
+  x = summarized_top_5_cause_data,
+  file = "outputs/data/summarized_top_5_cause_data.csv"
+)
 

### 2024-03-15 scripts/04-charts.R

@@ -15,16 +15,17 @@ library(knitr)
 library(janitor)
 library(ggplot2)
 library(tidyverse)
+library(tidybayes)
 
 
 
 #### load data ####
-filtered_data <- read_csv("outputs/data/filtered_data.csv")
+top_5_cause_data <- read_csv("outputs/data/top_5_cause_data.csv")
+summarized_top_5_cause_data <- read_csv("outputs/data/summarized_top_5_cause_data.csv")
 
 
-
-#### generate charts ####
-filtered_data |>
+#### top 5 causes ####
+top_5_cause_data |>
   ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
   geom_line() +
   theme_minimal() +
@@ -33,9 +34,27 @@ filtered_data |>
   facet_wrap(vars(cause), dir = "v", ncol = 1) +
   theme(legend.position = "none")
 
+
+
+#### summarized top 5 causes ####
+summarized_top_5_cause_data |>
+  ggplot(aes(x = calendar_year, y = sum_deaths)) +
+  geom_line() +
+  theme_minimal() +
+  scale_x_continuous(breaks = unique(summarized_top_5_cause_data$calendar_year)) +
+  labs(x = "Year", y = "Sum of Deaths") +
+  theme(legend.position = "none")
+
+
+
+#### poisson result ####
 pp_check(cause_of_death_alberta_poisson) +
   theme(legend.position = "bottom")
 
+
+
+#### negative binomial result ####
 pp_check(cause_of_death_alberta_neg_binomial) +
   theme(legend.position = "bottom")
+
 

### 2024-03-15 outputs/paper/paper.qmd
introduction updated

@@
 
-Alberta, a province known for its vibrant culture and dynamic economy, faces ongoing challenges in public health, with mortality rates reflecting a complex interplay of socioeconomic, environmental, and healthcare factors.
-This paper aims to delve into the intricacies of mortality rates in Alberta, exploring the trends, causes, and determinants that shape the province's public health landscape.
-By leveraging data on causes of death, including chronic diseases, accidents, and emerging health threats, we provide a detailed analysis of mortality patterns and their implications for Alberta's healthcare system and policy-making.
+# Introduction
+
+Alberta, often celebrated for its vibrant culture and robust economy, grapples with the complexities of public health challenges that reverberate across its diverse population.
+In the face of these challenges, mortality rates emerge as a telling metric, offering insights into the health outcomes shaped by an interwoven fabric of socioeconomic conditions, environmental factors, and healthcare system efficacy.
+The significance of these mortality rates transcends mere numbers; they mirror the collective well-being of the community and serve as a barometer for the effectiveness of public health policies and interventions.
+
+The exploration of mortality trends within this province is not merely an academic exercise; it is an imperative assessment in an era where healthcare systems worldwide are continually tested by emerging threats and chronic conditions alike.
+As such, this paper endeavors to dissect the evolving narrative of Alberta's mortality statistics, casting light on both long-standing and emerging causes of death.
+Our intent is not to pinpoint direct causality but rather to contextualize these trends within the broader panorama of Alberta's dynamic public health landscape.
+
+This endeavor goes beyond the superficial examination of statistics.
+It seeks to unravel the intricate tapestry of factors that dictate mortality trends, such as the shifting demographics of an aging population, the variable tide of health-related behaviors, and the silent encroachment of environmental changes.
+By synthesizing data on mortality causes, this paper aims to contribute a nuanced analysis that could potentially steer the province's healthcare policies towards more informed and strategic directions.
+It is within this framework that we lay the foundation for a quantitative assessment, poised to uncover the undercurrents that shape the health outcomes of Alberta's populace.
 

### 2024-03-15 outputs/paper/paper.qmd
intro updated
abstract added

@@
 date-format: long
-abstract: "Abstract"
+abstract: "This study is focusing on the mortality trends in Alberta, with an emphasis on understanding the broader public health implications.
+This study offers a panoramic view of the health challenges and demographic shifts influencing the mortality rates in the province.
+The findings highlight an increase in mortality over recent years, with implications that extend beyond individual health conditions to encompass wider societal and environmental factors.
+The paper underscores the importance of data-driven approaches in shaping effective public health strategies and policies."
 format: pdf
@@
 ```
-
-# Introduction
 

### 2024-03-15 outputs/paper/paper.qmd
final paper

--- 2024-03-15 outputs/paper/paper.qmd
@@
 date-format: long
-abstract: "Abstract"
+abstract: "This study focuses on the mortality trends in Alberta, emphasizing the broader public health implications.
+It offers a panoramic view of the health challenges and demographic shifts influencing the province's mortality rates.
+The findings highlight an increase in mortality in recent years, with implications extending beyond individual health conditions to encompass wider societal and environmental factors.
+This paper underscores the importance of data-driven approaches in shaping effective public health strategies and policies."
 format: pdf
@@
 # load packages
-packages <- c("tidyverse", "boot", "broom.mixed", "collapse", "dataverse", "gutenbergr", "janitor", "knitr", "marginaleffects", "modelsummary", "rstanarm", "tidybayes")
+packages <- c("tidyverse", "boot", "broom.mixed", "collapse", "dataverse", "gutenbergr", "janitor", "knitr", "marginaleffects", "modelsummary", "rstanarm", "tidybayes", "loo")
 
@@
 
-Alberta, a province known for its vibrant culture and dynamic economy, faces ongoing challenges in public health, with mortality rates reflecting a complex interplay of socioeconomic, environmental, and healthcare factors.
-This paper aims to delve into the intricacies of mortality rates in Alberta, exploring the trends, causes, and determinants that shape the province's public health landscape.
-By leveraging data on causes of death, including chronic diseases, accidents, and emerging health threats, we provide a detailed analysis of mortality patterns and their implications for Alberta's healthcare system and policy-making.
+Alberta, celebrated for its vibrant culture and robust economy, faces public health challenges that impact its diverse population.
+Mortality rates, as a key metric of economic level [@citeMortalityEconomy], influenced by a complex mix of socioeconomic conditions, environmental factors, and healthcare system performance.
+These rates reflect the community's collective well-being and gauge the effectiveness of public health policies and interventions.
+
+Exploring mortality trends in Alberta is crucial, given the global healthcare challenges posed by new threats and chronic conditions.
+This paper aims to analyze Alberta's mortality statistics, highlighting both longstanding and emerging causes of death.
+Our goal is to understand these trends within Alberta's dynamic public health context, rather than pinpoint direct causality.
+
+Our analysis delves into the factors influencing mortality rates, including demographic shifts, health-related behaviors [@citeBehavCause], government policies [@citeEconCause], and environmental changes [@citeEnvCause].
+By examining data on mortality causes, we aim to offer a nuanced perspective that could guide the province's healthcare policy towards more informed and strategic decisions.
+This study sets the stage for a quantitative assessment, seeking to reveal the underlying forces shaping health outcomes in Alberta.
+
+The subsequent sections of this paper are organized to facilitate a comprehensive understanding of the study and its implications.
+After this introduction, @sec-data outlines the methodology used in the analysis, emphasizing transparency and replicability.
+@sec-res presents the findings in detail, charting the trends in Alberta's mortality.
+@sec-dis discusses these findings from various perspectives as mentioned earlier.
+Finally, @sec-con concludes the paper, summarizing the key insights and findings of this research.
 
@@
 Data used in this paper was cleaned and processed with the programming language R [@citeR].
-Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary], `tidybayes` [@citeTidybayes].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary], `tidybayes` [@citeTidybayes], `loo` [@citeLoo].
 
@@
 
-In the pursuit of understanding and modeling mortality trends within Alberta, our study primarily utilizes the "Leading Causes of Death" dataset made available by the Alberta Open Government Portal.
-This comprehensive data source offers an extensive overview of mortality statistics across the province, categorized by various causes of death.
-The Alberta Open Government Portal, a key initiative for promoting transparency and public engagement, hosts this dataset at open.alberta.ca [@citeDatasource], providing open access to valuable health-related data.
+Our study delves into Alberta's mortality trends using the "Leading Causes of Death" dataset from the Alberta Open Government Portal [@citeDatasource].
+This dataset provides a comprehensive overview of mortality statistics throughout the province, organized by cause of death.
+As a cornerstone of transparency and public engagement, the Alberta Open Government Portal hosts this dataset at open.alberta.ca, offering open access to crucial health data.
 
-The "Leading Causes of Death" dataset encompasses a detailed compilation of death counts and mortality rates spanning multiple years.
-It includes categorizations by age, sex, and specific causes of death, ranging from chronic diseases to accidents and emergent health threats.
-This granularity allows for a nuanced analysis of mortality trends, facilitating the identification of prevalent health issues and potential public health interventions.
+The dataset includes detailed death counts and mortality rates over several years, with breakdowns by age, sex, and specific causes, from chronic diseases to accidents and new health threats.
+This detail enables a nuanced analysis of mortality trends and the identification of key health challenges, guiding potential public health interventions.
 
-The selection of this data source was driven by its credibility, comprehensiveness, and relevance to our research objectives.
-Being an official government resource, the dataset ensures reliability and accuracy, crucial for the validity of our analytical models.
-The dataset's extensive coverage of demographic and cause-specific mortality information enables a thorough exploration of the factors influencing mortality rates in Alberta.
-Furthermore, the data's organization and accessibility significantly aid in the efficient processing and modeling of mortality trends.
+We selected this dataset for its reliability, depth, and alignment with our study's goals.
+As an official government source, it provides accurate and trustworthy data essential for our analysis.
+Its comprehensive demographic and cause-specific details allow for an in-depth examination of mortality influences in Alberta.
+Additionally, the dataset's structure supports efficient analysis and modeling of mortality trends.
 
@@
 
-# Results
+The choice of models for our analysis was the Poisson and Negative Binomial.
+These models are particularly suited for mortality data analysis, aligning well with the characteristics of count data.
+Mortality counts, which represent the number of deaths over specific periods, naturally follow the discrete distribution assumed by the Poisson model.
+This model assumes equal mean and variance, a condition that typically applies to independent events occurring at a constant rate.
+The Poisson model is valued for the interpretability of its parameters; its exponential coefficients reflect the change in mortality rate per unit change in a predictor, providing clear insights into factors influencing mortality rates.
+This clarity is crucial for deriving actionable insights from the data.
+However, mortality data often exhibit overdispersion—where the variance exceeds the mean—challenging the Poisson model’s assumption.
+The Negative Binomial model addresses this by introducing an extra parameter to handle overdispersion, making it a more adaptable choice for analyzing data with greater variability.
+
+# Results {#sec-res}
 
@@
 #| fig-pos: 'h'
-#| label: fig-trends
-#| fig-cap: Trend for Alberta Mortality
-#| fig-subcap: ["Top 5 Causes","Sum of Causes"]
-#| layout-ncol: 2
-#| layout-nrow: 1
+#| label: fig-trends-1
+#| fig-cap: Trend for Alberta Mortality-Top 5 Causes
 
@@
         axis.text.y = element_text(size = rel(0.7)))
+
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-trends-2
+#| fig-cap: Trend for Alberta Mortality-Sum of Causes
 
@@
 
+\newpage
+
 For the Poisson model, the expected log count is defined as:
@@
 
-Then to see how good two models fits the original data.
-We can reffer to the values in the lower half of @tbl-summary, but it's always more clear to see the visual, as depicted by the graphs for the Poisson (@fig-models-1) and negative binomial models (@fig-models-2), the diagnostic plots further elucidate the model fits.
-The negative binomial model, appears to capture the dispersion in the data more effectively than the Poisson model, which is evident in two plots.
-The spread of the simulated lines (y_rep) around the observed data (y) suggests that the negative binomial model may accommodate the overdispersion present in the data, which is a common characteristic in count data such as mortality rates.
+Then, to see how well the two models fit the original data, we can refer to the values in the lower half of @tbl-summary.
+However, it's always clearer to see this visually, as depicted by the graphs for the Poisson (@fig-models-1) and Negative Binomial models (@fig-models-2).
+The diagnostic plots further elucidate the fits of the models.
+The spread of the simulated lines (y_rep) around the predicted data (y) suggests that the Negative Binomial model may better accommodate the overdispersion present in the data, which is a common characteristic of count data such as mortality rates.
 
-The presence of multiple peaks in the second plot's simulated data distributions indicates potential subgroups within the data or non-linearity in the mortality trends that the Poisson model may not be accounting for.
+The presence of multiple peaks in the simulated data distributions of @fig-models-2 indicates potential subgroups within the data or non-linearity in the mortality trends that the Poisson model may not be accounting for.
 Such patterns underscore the importance of considering different model structures when dealing with complex datasets.
-These graphical diagnostics not only affirm the adequacy of the negative binomial model over the Poisson model for this particular dataset but also highlight the necessity of thorough model checking in the process of statistical analysis.
+The Negative Binomial model appears to capture the dispersion in the data more effectively than the Poisson model.
+This not only evident in the plots, but also by the larger ELPD (Expected Log Pointwise Predictive Density) value showing in @tbl-cross_validation, as a higher ELPD indicates a model with better predictive accuracy.
+Furthermore, this better prediction by the Negative Binomial model was consistent with the predictions for the two models mentioned in the method section of @sec-data.
 
-# Discussion
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: tbl-cross_validation
+#| tbl-cap: cross validation of modeling
+
+options(mc.cores = 2)
+
+poisson <- loo(cause_of_death_alberta_poisson, cores = 2)
+neg_binomial <- loo(cause_of_death_alberta_neg_binomial, cores = 2)
+
+loo_compare(poisson, neg_binomial)
+```
+
+# Discussion {#sec-dis}
 
 The preceding sections have laid the groundwork for a comprehensive analysis of mortality trends in Alberta, paving the way for this discussion.
-The application of Poisson and Negative Binomial models to the data has illuminated key trends and has provided quantitative support to the qualitative observations made earlier.
-The positive coefficients for "other chronic" and "neoplasms" in both models suggest that deaths from these causes have steadily increased, raising questions about the broader factors at play beyond the healthcare system's direct influence.
+
+## Demographic Shifts
+
+Alberta's demographic landscape has been undergoing significant transformations, predominantly fueled by a steady influx of immigrants from within Canada and across the globe [@citeAlbertaDemog].
+This diversification of the population not only enriches the province's cultural fabric but also introduces complex dynamics in public health management.
+Immigrants, adapting to a new environment, might face unique health challenges including climate-related discomfort and exposure to diseases not prevalent in their countries of origin.
+These factors, coupled with the stressors associated with resettlement, could potentially influence mortality rates in ways that are distinct from the native population.
+
+The demographic shift towards a more diverse population composition also brings into focus the aging demographic.
+As the overall population grows, so does the proportion of elderly residents, a group inherently at higher risk of mortality due to chronic diseases and other age-related health issues.
+This aging demographic, compounded by the province's overall population growth, naturally leads to an increase in the number of deaths.
+However, interpreting these rising death tolls requires careful consideration of the expanding population base [@citeAlbertaPop]; a larger population inherently includes more individuals at risk of mortality, thereby elevating the absolute number of deaths even if the mortality rate (deaths per capita) remains constant or decreases.
+
+## Health-related Behaviors
+
+Health-related behaviors, encompassing lifestyle choices such as diet, physical activity, smoking, and alcohol consumption, significantly influence mortality trends in Alberta.
+These behaviors, often rooted in socio-economic conditions, cultural norms, and individual psychology, have a profound impact on the prevalence and progression of chronic diseases, including heart disease, diabetes, and various forms of cancer.
+For instance, the high prevalence of sedentary lifestyles coupled with increasing consumption of processed foods contributes to obesity and related health complications, which can elevate mortality rates.
+Smoking and excessive alcohol consumption continue to be major risk factors for respiratory diseases [@citeBehavCause], liver conditions, and cancers, despite public health campaigns aimed at curbing these behaviors.
+
+Refer back to @tbl-summary, the positive coefficients for "other chronic" and "neoplasms" in both models suggest that deaths from these causes have steadily increased, raising questions about the broader factors at play beyond the healthcare system's direct influence.
 Simultaneously, the negative coefficient for "obstructive pulmonary" diseases suggests a decline in mortality from these conditions, potentially reflecting successful public health initiatives and changing social habits, such as reduced smoking rates.
 
-However, the increasing overall mortality trend signals a need to consider demographic shifts, such as population growth and aging, as significant contributors to the observed patterns.
-The spike in deaths from 2020 onwards aligns with the onset of the COVID-19 pandemic, underscoring the substantial impact of infectious disease outbreaks on mortality statistics.
-This outlier emphasizes the importance of having robust health surveillance systems that can adapt to sudden changes in public health landscapes.
+## Government Policies
 
-Furthermore, the diagnostic plots from both models underscore the necessity for careful selection of appropriate statistical methods when modeling complex health data.
-The better fit of the Negative Binomial model compared to the Poisson model, as indicated by the diagnostic checks, suggests that overdispersion—a common feature in count data—must be accounted for to avoid underestimating the variability in the data.
+Government policies, particularly those related to austerity measures, play a significant role in shaping public health outcomes and mortality rates [@citeEconCause].
+Austerity policies, which are often implemented to reduce government debt through spending cuts and increased taxes, can have far-reaching effects on the social determinants of health.
+For instance, reductions in funding for public healthcare services can limit access to preventive care, screenings, and treatment for chronic diseases, potentially leading to an increase in mortality rates.
+Additionally, austerity measures that result in cuts to social services can exacerbate poverty, unemployment, and housing instability, further impacting individuals' health and well-being.
 
-The limitations of the current study also warrant discussion.
-While the models applied provide valuable insights, they do not capture all the nuances of the data.
-Future research could explore the inclusion of additional variables, such as economic indicators, healthcare access measures, and lifestyle factors, to better understand the determinants of mortality.
-Moreover, the investigation of interactions between causes of death and demographic variables could yield a more granular understanding of the dynamics at play.
+In Alberta, where public health policy can significantly influence healthcare delivery and social welfare programs, the implementation of austerity measures may pose challenges to achieving optimal health outcomes.
+Although there was no implimentation of austerity policies recent years, some past implementation has already proven to have a significant negative impact on society [@citeAusterity].
+The relationship between such policies and mortality underscores the need for careful consideration of the health impacts of fiscal decisions.
+Policymakers must balance budgetary constraints with the imperative to protect and enhance public health, ensuring that austerity measures do not disproportionately harm vulnerable populations or erode the quality of healthcare services.
 
-Ultimately, this study serves as a call to action for ongoing surveillance, research, and a proactive approach to public health planning.
-By continuing to refine our models and expand our datasets, we can better anticipate and mitigate the factors contributing to mortality, striving for a healthier future for all Albertans.
+## Environmental Changes
 
-# Conclusion
+Environmental changes in Alberta, ranging from air and water pollution to climate variability and extreme weather events, significantly influence public health and mortality trends.
+The province, known for its diverse landscapes and economic reliance on industries such as oil and gas, agriculture, and forestry, faces unique environmental challenges that directly impact the health of its residents.
+Air pollution [@citeAirPollu]
+, resulting from industrial activities and urban traffic, is linked to respiratory diseases, cardiovascular conditions, and increased mortality rates.
+Industrial waste issues [@citeFactoryPollu], such as sand and water, can leading to health complications and diseases.
+Besides, even seemingly insignificant temperature changes in living environments can have an impact on human health [@citeEnvCause], and such changes are mainly attributed to global warming.
 
-In conclusion, our exploration of mortality trends in Alberta has provided a statistical narration of the province's public health status over the past two decades.
-We have identified critical causes of death and trends that warrant attention from both healthcare providers and policymakers.
-The analysis has also emphasized the significance of employing appropriate statistical models to accurately represent and understand complex health data.
-As we move forward, it is imperative to integrate these findings into a strategic framework that enhances Alberta's health system responsiveness and effectiveness in the face of demographic and epidemiological shifts.
-Through vigilance and informed action, we can aim to influence these mortality trends positively, enhancing the health and well-being of the Alberta population.
+# Conclusion {#sec-con}
+
+In conclusion, our examination of Alberta's mortality trends over the last two decades has shed light on the province's public health landscape.
+We've pinpointed key causes of death and trends that demand attention from healthcare providers and policymakers.
+Our work underscores the necessity of using suitable statistical models to grasp complex health data accurately.
+
+Our study reveals the diverse factors affecting public health in Alberta, including demographic changes, health behaviors, government policies, and environmental challenges.
+The influx of immigrants has significantly shifted Alberta's demographic makeup, bringing new health issues and diversifying healthcare needs.
+Lifestyle factors, such as diet, physical activity, and substance use, significantly influence health outcomes, highlighting the importance of targeted public health initiatives.
+Moreover, government policies and environmental factors, like pollution and climate change, present ongoing challenges to public health.
 

### 2024-03-15 outputs/paper/paper.qmd
paper final draft

@@
 ---
-title: "Modeling Mortality Trends in Alberta"
-subtitle: "A Quantitative Approach to Public Health Dynamics"
+title: "Deciphering Mortality in Alberta: A Multifaceted Analysis"
+subtitle: "A Quantitative Approach on Demographics, Behaviors, Policies, and Environment"
 author: 

### 2024-03-16 outputs/paper/paper.qmd
final draft for scripts

@@
 
+## Possible Improvements
+
+While our study sheds light on Alberta's mortality trends, it's important to recognize its limitations.
+Our analysis depends on available data, which might not capture every detail affecting mortality rates.
+The methods we used, although useful for spotting general trends, might oversimplify the complex factors at play, including how policies or environmental changes impact health.
+Future research should aim to use more detailed data and explore these complexities more deeply.
+It's also essential to look at the long-term effects of changes in public health policies and the environment.
+By digging deeper into these areas and using a mix of quantitative and qualitative research, we can better understand and address the factors that influence health outcomes in Alberta.
+
 # Conclusion {#sec-con}

### 2024-03-16 scripts/00-simulate_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: 
+# Purpose: This code creates simulated mortality data by cause over years, and visualizes this data with a ggplot chart that includes points for each data entry and a linear trend line, all within a minimalistic theme.
 # Author: Ziheng Zhong
 # Date: 10 March 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
@@ -35,5 +35,3 @@ ggplot(simulation_data, aes(x = year, y = deaths, color = cause)) +
        x = "Year",
        y = "Number of Deaths") +
   scale_x_continuous(breaks = unique(simulation_data$year))
-
-

### 2024-03-16 scripts/01-data_downloading.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: 
+# Purpose: This code downloads and processes mortality data from an online source by specifying column types and skipping initial rows, and then saves the cleaned dataset for future use.
 # Author: Ziheng Zhong
 # Date: 10 March 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
@@ -19,8 +19,8 @@ library(tidyverse)
 raw_data <-
   read_csv(
     "https://open.alberta.ca/dataset/03339dc5-fb51-4552-97c7-853688fc428d/resource/3e241965-fee3-400e-9652-07cfbf0c0bda/download/deaths-leading-causes.csv",
-    skip = 2,
-    col_types = cols(
+    skip = 2, # Skip the first two rows (header) of the CSV file
+    col_types = cols( # Defining column data types
       `Calendar Year` = col_integer(),
       Cause = col_character(),
       Ranking = col_integer(),
@@ -35,5 +35,3 @@ write_csv(
   x = raw_data,
   file = "inputs/data/raw_data.csv"
 )
-
-

### 2024-03-16 scripts/02-data_cleaning.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: 
+# Purpose: This code performs data cleaning and aggregation tasks, identifies and summarizes the top 5 causes of death in Alberta for 2019, and saves the processed datasets for further analysis or reporting.
 # Author: Ziheng Zhong
 # Date: 10 March 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
@@ -23,48 +23,50 @@ raw_data <- read_csv("inputs/data/raw_data.csv")
 
 
 #### data cleaning ####
-cleaned_data <- raw_data %>%
-  clean_names() %>%
-  add_count(cause) %>%
-  mutate(cause = str_trunc(cause, 30))
+cleaned_data <- raw_data %>%  # Assigns the result of the following operations to `cleaned_data`
+  clean_names() %>%  # Converts column names of `raw_data` to a consistent, clean format
+  add_count(cause) %>%  # Adds a new column that counts the number of occurrences for each `cause`
+  mutate(cause = str_trunc(cause, 30))  # Truncates the `cause` column strings to 30 characters, for beter readability
 
 
 
 #### check top 8 major causes in 2019 ####
-cleaned_data |>
+cleaned_data |>  # Starts with the `cleaned_data` dataset
   filter(
-    calendar_year == 2019,
-    ranking <= 8
-  ) |>
-  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>
+    calendar_year == 2019,  # Keeps only the rows where `calendar_year` is 2019
+    ranking <= 8  # Keeps only the rows where `ranking` is 8 or less
+  ) |> 
+  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>  # Formats `total_deaths` column, adding commas as thousands separators
   kable(
-    col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
-    align = c("l", "r", "r", "r", "r"),
-    digits = 0, booktabs = TRUE, linesep = ""
+    col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),  # Specifies column names for the table
+    align = c("l", "r", "r", "r", "r"),  # Sets alignment for each column: left for the first and right for the others
+    digits = 0,  # Specifies that numbers should be displayed with no decimal places
+    booktabs = TRUE,  # Uses booktabs style for the table, providing a more professional look
[…]

### 2024-03-16 scripts/03-modeling.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: 
+# Purpose: This code fits Poisson and Negative Binomial models to analyze Alberta's mortality rates, and summarizes the modeling results with customized naming for clarity.
 # Author: Ziheng Zhong
 # Date: 10 March 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
@@ -16,34 +16,34 @@ library(modelsummary)
 
 
 #### load data ####
-filtered_data <- read_csv("outputs/data/filtered_data.csv.csv")
+top_5_cause_data <- read_csv("outputs/data/top_5_cause_data.csv")
 
 
 
 #### poisson model ####
-cause_of_death_alberta_poisson <-
-  stan_glm(
-    total_deaths ~ cause,
-    data = filtered_data,
-    family = poisson(link = "log"),
-    seed = 123
+cause_of_death_alberta_poisson <-  # Assigns the result of the following model fitting operation to `cause_of_death_alberta_poisson`
+  stan_glm(  # Calls the `stan_glm` function to fit a generalized linear model using Stan
+    total_deaths ~ cause,  # Defines the model formula: `total_deaths` as the response variable, with `cause` as the predictor
+    data = top_5_cause_data,  # Specifies the dataset to use for the model fitting
+    family = poisson(link = "log"),  # Sets the model family to Poisson, indicating the response variable is count data, and specifies the link function as logarithmic
+    seed = 123  # Sets a seed for random number generation to ensure reproducibility of the model fitting process
   )
 
 
 
 #### negative binomial model ####
-cause_of_death_alberta_neg_binomial <-
-  stan_glm(
-    total_deaths ~ cause,
-    data = filtered_data,
-    family = neg_binomial_2(link = "log"),
-    seed = 123
+cause_of_death_alberta_neg_binomial <-  # Assigns the result of the model fitting to `cause_of_death_alberta_neg_binomial`
+  stan_glm(  # Calls the `stan_glm` function to fit a generalized linear model using Stan
+    total_deaths ~ cause,  # Defines the model formula: `total_deaths` as the response variable, with `cause` as the predictor
+    data = top_5_cause_data,  # Specifies the dataset to use for the model fitting
+    family = neg_binomial_2(link = "log"),  # Sets the model family to Negative Binomial (version 2), suitable for count data with overdispersion, with a logarithmic link function
[…]

### 2024-03-16 scripts/04-charts.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: 
+# Purpose: This code analyzes and visualizes Alberta's top 5 causes of mortality, assesses model fits for Poisson and Negative Binomial distributions, and sets up the workspace with essential libraries and data.
 # Author: Ziheng Zhong
 # Date: 10 March 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
@@ -25,36 +25,34 @@ summarized_top_5_cause_data <- read_csv("outputs/data/summarized_top_5_cause_dat
 
 
 #### top 5 causes ####
-top_5_cause_data |>
-  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
-  geom_line() +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  labs(x = "Year", y = "Annual number of deaths in Alberta") +
-  facet_wrap(vars(cause), dir = "v", ncol = 1) +
-  theme(legend.position = "none")
+top_5_cause_data |>  # Uses the `top_5_cause_data` dataset
+  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +  # Initializes a ggplot with `calendar_year` on the x-axis, `total_deaths` on the y-axis, and colors lines by `cause`
+  geom_line() +  # Adds line geometries to plot the trends of deaths over years for each cause
+  theme_minimal() +  # Applies a minimal theme to the plot for a clean and modern appearance
+  scale_color_brewer(palette = "Set1") +  # Uses a color palette from Color Brewer for the line colors, enhancing visual distinction
+  labs(x = "Year", y = "Annual number of deaths in Alberta") +  # Labels the x-axis as "Year" and the y-axis as "Annual number of deaths in Alberta"
+  facet_wrap(vars(cause), dir = "v", ncol = 1) +  # Creates separate plots (facets) for each cause, arranged vertically with one column
+  theme(legend.position = "none")  # Hides the legend since the facet titles already indicate the cause
 
 
 
 #### summarized top 5 causes ####
-summarized_top_5_cause_data |>
-  ggplot(aes(x = calendar_year, y = sum_deaths)) +
-  geom_line() +
-  theme_minimal() +
-  scale_x_continuous(breaks = unique(summarized_top_5_cause_data$calendar_year)) +
-  labs(x = "Year", y = "Sum of Deaths") +
-  theme(legend.position = "none")
+summarized_top_5_cause_data |>  # Uses the `summarized_top_5_cause_data` dataset
[…]

### 2024-03-16 scripts/05-tests.R

@@ -1,0 +1,98 @@
+#### Preamble ####
+# Purpose: The code defines a test suite for validating data integrity, including checks for positive calendar years, four-digit year format, non-empty cause entries, and non-negative death counts across multiple datasets within a research project.
+# Author: Ziheng Zhong
+# Date: 10 March 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+
+
+
+#### Workspace setup ####
+# setup all libraries
+library(readr)
+library(testthat)
+library(tidyverse)
+
+
+
+#### load data ####
+setwd("/Users/ijustin/Library/CloudStorage/Dropbox/Code/STA302/Research_Papers/Mortality_Rate_in_Alberta")
+raw_data <- readr::read_csv("inputs/data/raw_data.csv")
+top_5_cause_data <- read_csv("outputs/data/top_5_cause_data.csv")
+summarized_top_5_cause_data <- read_csv("outputs/data/summarized_top_5_cause_data.csv")
+
+
+
+#### raw_data testing ####
+test_that("Calendar Year values are positive", {
+  # Checks that all values in the `Calendar Year` column are non-negative, excluding NA values
+  expect_true(all(raw_data$`Calendar Year` >= 0, na.rm = TRUE))
+})
+
+test_that("Calendar Year values have four digits", {
+  # Checks that all values in the `Calendar Year` column have four digits, excluding NA values
+  expect_true(all(nchar(as.character(raw_data$`Calendar Year`)) == 4, na.rm = TRUE))
+})
+
+test_that("Cause column contains no empty values", {
+  # Checks that the `Cause` column contains no NA or empty strings
+  expect_true(all(!is.na(raw_data$Cause) & raw_data$Cause != ""))
+})
+
+test_that("Deaths column contains no negative values", {
+  # Checks that all values in the `Total Deaths` column are non-negative, excluding NA values
+  expect_true(all(raw_data$`Total Deaths` >= 0, na.rm = TRUE))
+})
+
+test_that("Ranking values are positive", {
+  # Checks that all values in the `Ranking` column are positive integers, excluding NA values
+  expect_true(all(raw_data$`Ranking` > 0, na.rm = TRUE))
+})
+
+
+
+#### top_5_cause_data testing ####
+test_that("Calendar Year values are positive", {
[…]

### 2024-03-16 outputs/paper/paper.qmd
paper final draft

@@
 
-Data used in this paper was cleaned and processed with the programming language R [@citeR].
-Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary], `tidybayes` [@citeTidybayes], `loo` [@citeLoo].
+Data used in this paper was cleaned, processed and tested with the programming language R [@citeR].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModelsummary], `tidybayes` [@citeTidybayes], `loo` [@citeLoo], `testthat` [@citeTestthat].
 
